---
title: "SQL Tutorial"
author: "Amanda Ng"

format:
  live-html:
    toc: true
    number-sections: false
  
engine: knitr

resources:
  - data


webr:
  packages:
    - tidyverse
    - DBI
    - duckdb

live:
  show-hints: true
  show-solutions: true
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) 
library(DBI) 
library(duckdb) 
library(tidyverse) 
# 1. Read your CSV 
mydata <- read.csv("data/nhanes.csv") 
nhanes_demographics <- read.csv("data/nhanes_demographics.csv") 
nhanes_health <- read.csv("data/nhanes_health.csv") 
nhanes_set1<- read.csv("data/nhanes_set1.csv") 
nhanes_set2<- read.csv("data/nhanes_set2.csv") 

# 2. Create an in-memory SQLite database 
con <- dbConnect(duckdb::duckdb(), dbdir = ":memory:")

# 3. Write the CSV into a SQL table 
dbWriteTable(con, "nhanes", mydata) 
dbWriteTable(con, "nhanes_demographics", nhanes_demographics) 
dbWriteTable(con, "nhanes_health", nhanes_health) 
dbWriteTable(con, "nhanes_set1", nhanes_set1)
dbWriteTable(con, "nhanes_set2", nhanes_set2)
```

## Basic query

### SELECT, FROM, LIMIT

A typical starting structure when using SQL is `SELECT variable FROM table`. This query will give a table showing only the selected variables from the table. To select multiple variables, we can use a comma `,` to separate them. The shortcut for showing all variables is `*`.

To avoid overwhelming the reader, we can limit the output table to show only the top observations using `LIMIT` after the `FROM` clause. For example:

::: {.panel-tabset}
# SQL

```{sql,connection=con}
SELECT Age, Education, Triglyceride
FROM nhanes
LIMIT 5
```

# R

```{r}
#| edit: false
mydata %>%
  select(Age, Education, Triglyceride)%>%
  head(5)
```
:::

We can also specify to show only distinct rows with `DISTINCT` in the `SELECT` rows. However, you are only allowed to distinct on a single variable or `DISTINCT *`, such as

::: {.panel-tabset}
# SQL

```{sql,connection=con}
SELECT DISTINCT Age
FROM nhanes
LIMIT 10;
```

# R

```{r}
#| edit: false
mydata %>%
  select(Age)%>%
  head(10)%>%
  distinct()
```
:::

To show distinct rows across multiple selected variables, we need to perform a `GROUP BY`, which will be introduced later.

### Create new variables with SELECT, AS

We can also rename the variable name using `AS`, such as

::: {.panel-tabset}
# SQL

```{sql,connection=con}
SELECT Age, Education AS Edu, Triglyceride
FROM nhanes
LIMIT 5
```

# R

```{r}
#| edit: false
mydata %>%
  select(Age, Edu = Education, Triglyceride) %>%
  head(5)
```
:::

To define new variables, we also do it within the `SELECT` clause such as

::: {.panel-tabset}
# SQL

```{sql,connection=con}
SELECT Age, Age + 10 AS age_10_years_later
FROM nhanes
LIMIT 5
```

# R

```{r}
#| edit: false
mydata %>%
  mutate(age_10_years_later = Age + 10)%>%
  select(Age, age_10_years_later)%>%
  head(5)
```
:::

We can also define a constant variable across rows.

::: {.panel-tabset}
# SQL

```{sql,connection=con}
SELECT 2026 AS constant_year
FROM nhanes
LIMIT 5
```

# R

```{r}
#| edit: false
mydata %>%
  mutate(constant_year = 2026)%>% 
  select(constant_year)%>%
  head(5)
```
:::

We can also concatenate strings using `||`.

::: {.panel-tabset}
# SQL

```{sql,connection=con}
SELECT Ethnicity||Marital_status AS combined_e_m
FROM nhanes
LIMIT 5
```

# R

```{r}
#| edit: false
mydata %>%
  mutate(combined_e_m = str_c(Ethnicity, Marital_status))%>% 
  select(combined_e_m)%>%
  head(5)
```
:::

### Conditional assignments with CASE WHEN

To assign value to a new variable based on some conditions, we use `CASE WHEN... THEN... ELSE...END`.

In this assignment, we transformed the coded `Gender` column into the text form column `Gender_name` where we assign the value `Male` when it is coded as 1, `Female` when it is coded as 2, and `Unknown` otherwise.

::: {.panel-tabset}
# SQL

```{sql,connection=con}
SELECT Gender,
      CASE WHEN Gender = 1 THEN 'Male'
          WHEN Gender = 2 THEN 'Female'
            ELSE 'Unknown'
        END AS Gender_name      
FROM nhanes
LIMIT 10
```

# R

```{r}
#| edit: false
mydata %>%
  mutate(Gender_name = case_when(Gender == 1 ~ 'Male', 
                                 Gender == 2 ~ 'Female',
                                 TRUE ~ 'Unknown'))%>% 
  select(Gender, Gender_name)%>%
  head(10)
```
:::

### Filtering rows with WHERE

Similar to `dplyr` in R, we can also filter to restrict our output to only show observations that satisfy certain conditions using `WHERE` clause. Note that the `WHERE` clause always comes after `FROM`. To build condition statements, we can use

-   Logical operators: `=`, `<`, `>`, `<=`, `>=`, and `!=`.
-   Boolean operators: `AND`, `OR`, and `NOT`

In this example, we show the top 5 observations with age greater than 20.

::: {.panel-tabset}
# SQL

```{sql,connection=con}
SELECT *
FROM nhanes
WHERE Age > 20
LIMIT 5
```

# R

```{r}
#| edit: false
mydata %>%
  filter(Age > 20)%>%
  head(5)
```
:::

SQL provides the `LIKE` operator to compare a string to a pattern. The pattern is a quoted string and can include these special characters:

-   `_` matches any single character
-   `%` matches any sequence of zero or more characters

This outputs all distinct `Ethnicity` which contains `Hispanic` in the entry.

::: {.panel-tabset}
# SQL

```{sql,connection=con}
SELECT DISTINCT Ethnicity
FROM nhanes
WHERE Ethnicity LIKE '%Hispanic%'
```

# R

```{r}
#| edit: false
mydata %>%
  filter(str_like(Ethnicity, '%Hispanic%'))%>%
  select(Ethnicity)%>%
  distinct()
```
:::

### ORDER BY

Using `ORDER BY`, we can order the rows in the table based on column(s) of your choice. By default, it orders the row in ascending order (`ASC`). If you want to show the row corresponding to the larger values as the top row, put `DESC` after your column name.

This outputs the top 5 distinct `Age` in descending order, the first row corresponds to the observation with largest age in the dataset.

::: {.panel-tabset}
# SQL

```{sql,connection=con}
SELECT DISTINCT Age
FROM nhanes
ORDER BY Age DESC
LIMIT 5
```

# R

```{r}
#| edit: false
mydata %>%
  select(Age)%>%
  arrange(desc(Age))%>%
  distinct()%>%
  head(5)
```
:::

## Grouping and Aggregation

Sometimes, we are interested in the group level statistical summaries. We can use `GROUP BY` and aggregation functions to so do.

Here are some common aggregation functions in SQL:

-   `AVG()`
-   `MAX()`
-   `MIN()`
-   `COUNT()`
-   `SUM()`

Tips:

-   use `COUNT(*)` is you want to summarize number of rows in the table
-   use `COUNT(DISTINCT variable)` is you want to summarize number of distinct values in a variable

For example, this shows the mean age for each education groups.

::: {.panel-tabset}
# SQL

```{sql,connection=con}
SELECT Education, AVG(Age) AS avg_age
FROM nhanes
GROUP BY Education
```

# R

```{r}
#| edit: false
mydata %>%
  group_by(Education)%>%
  summarize(avg_age = mean(Age))
```
:::

Note that, when we use `GROUP BY`, we must include the grouping column in the `SELECT` clause. Any columns storing individual-level values should not be included in `SELECT` after grouping since we are collapsing the dataset into group level observations. In other words, you can only include the grouping column and aggregated measures applied on the grouping column.

You can also group by more than one column, such as:

::: {.panel-tabset}
# SQL

```{sql,connection=con}
SELECT Education, Gender, AVG(Age) AS avg_age
FROM nhanes
GROUP BY Education, Gender
```

# R

```{r}
#| edit: false
mydata %>%
  group_by(Education, Gender) %>%
  summarise(avg_age = mean(Age, na.rm = TRUE))
```
:::

When we use `GROUP BY`, we are defining the units of aggregation. A common misconception is that grouping by multiple columns creates separate groups for each column independently. In reality, grouping uses the **combination** of the columns to define each group. In this example,

- Grouping by one column (e.g., `Education`) creates one group for each unique Education level.
- Grouping by two columns (e.g., `Education` and `Gender`) creates one group for each **unique pair** of values.

As a result, the same Education level can appear in multiple rows because each row represents a different combination with the second grouping variable. For example:

- `Some college or AA degree` + `1`
- `Some college or AA degree` + `2`

both belong to the same `Education` category, but they are different groups because the `Gender` value changes.

## Filter groups with HAVING

Similar to `WHERE`, `HAVING` allows us to filter rows in the table after the `GROUP BY` clause. Using `WHERE` after `GROUP BY` is invalid.

For instance, we only want the average age of observations with `College graduate or above` education.

::: {.panel-tabset}
# SQL

```{sql,connection=con}
SELECT Education, AVG(Age) AS avg_age
FROM nhanes
GROUP BY Education
HAVING Education = 'College graduate or above'
```

# R

```{r}
#| edit: false
mydata %>%
  group_by(Education)%>%
  filter(Education == 'College graduate or above')%>%
  summarize(avg_age = mean(Age))
```
:::

If you want to filter rows such that the variable value is in a list of values, we can use `IN` and `(your_list_of_values)`.

For instance, we only want to calculate the average ages of observations with `College graduate or above`, `Some college or AA degree`, and `9-11th grade` respectively.


::: {.panel-tabset}
# SQL
```{sql,connection=con}
SELECT Education, AVG(Age) AS avg_age
FROM nhanes
GROUP BY Education
HAVING Education IN ('College graduate or above', 'Some college or AA degree', '9-11th grade')
```

# R

```{r}
#| edit: false
mydata %>%
  group_by(Education) %>%
  summarise(avg_age = mean(Age, na.rm = TRUE)) %>%
  filter(Education %in% c(
    "College graduate or above",
    "Some college or AA degree",
    "9-11th grade"
  ))
```
:::

## Set operations on row level

SQL provides several set operations that allow you to combine results of two queries (or two datasets) on row level.

Note that an operand to a set operator must be a complete query. For instance, if we wanted to conduct `SET_OPERATION` on tables `nhanes_set1` and `nhanes_set2`, we couldn’t just write `(nhanes_set1) SET_OPERATION (nhanes_set2)`.

### UNION

`UNION` returns all rows that appear in either of the two result sets. In the example below, we have `nhanes_set1` containing information from participants with ID 93731, 93801,and 93823; and `nhanes_set2` containing information from participant with ID 93823, 93830, and 93840. The result outputs 5 rows in total. Notice that participant 93823 appears twice since it exists on both `nhanes_set1` and `nhanes_set2`. `UNION` automatically eliminates duplicates.

::: {.panel-tabset}
# SQL

```{sql,connection=con}
SELECT *
FROM nhanes_set1
UNION 
SELECT * 
FROM nhanes_set2
```

# R

```{r}
#| edit: false
union(nhanes_set1, nhanes_set2)
```
:::

To keep the duplicates, we can use `UNION ALL`. The result outputs 6 rows in total.

::: {.panel-tabset}
# SQL

```{sql,connection=con}
SELECT *
FROM nhanes_set1
UNION ALL 
SELECT * 
FROM nhanes_set2
```

# R

```{r}
#| edit: false
bind_rows(nhanes_set1, nhanes_set2)
```
:::

### INTERSECTION

`INTERSECT` returns the rows that appear in both sets. Using the same `nhanes_set1` and `nhanes_set2`, the result only outputs the row corresponding to participant 93823.

::: {.panel-tabset}
# SQL

```{sql,connection=con}
SELECT *
FROM nhanes_set1
INTERSECT 
SELECT * 
FROM nhanes_set2
```

# R

```{r}
#| edit: false
intersect(nhanes_set1, nhanes_set2)
```
:::

### EXCEPT

`EXCEPT` returns rows from the first set (the one on the `FROM` query) that do not appear in the second.

::: {.panel-tabset}
# SQL

```{sql,connection=con}
SELECT *
FROM nhanes_set1
EXCEPT 
SELECT * 
FROM nhanes_set2
```

# R

```{r}
#| edit: false
setdiff(nhanes_set1, nhanes_set2)
```
:::

Note that `EXCEPT` removes all occurrences of duplicate data from the first set.

If you wish to removes one occurrence of duplicate data from the first set for every occurrence in the second set, use `EXCEPT ALL`.

## Merging datasets on column level

We can merge datasets using `JOIN` if they share common columns. The general syntax is `A JOIN_TYPE B`, where `JOIN_TYPE` depends on the join operation we want to perform. This syntax would perform the join based on all shared attributes (columns) between A and B. If the shared attributes you wish to merge on are named differently in A and B, we can use `ON` to specify the condition. For instance, `A JOIN_TYPE B ON A.colname1 = B.colname2` will join rows from A and B by matching values between `colname1` from A and `colname2` from B.

In this section, we will use the following datasets.

`nhanes_demographics` includes ID, Age, Gender and Education level of 5 patients.

```{sql,connection=con}
SELECT *
FROM nhanes_demographics
```

`nhanes_health` includes ID, Heart attack status, and Stroke Status of 6 patients.

```{sql,connection=con}
SELECT *
FROM nhanes_health
```

Note that ID 93731, 93840, and 93920 appear on both table. ID 93972 and 94069 only appear in `nhanes_demographics` and ID 102838, 102880, and 102947 only appear in `nhanes_health`.

Suppose we are only interested in patients' age and their stroke status, there are 4 main types of joins:

### Inner Join

An inner natural join:

::: {.panel-tabset}
# SQL

```{sql,connection=con}
SELECT SEQN, Age, Stroke
FROM nhanes_demographics
NATURAL JOIN nhanes_health
```

# R

```{r}
#| edit: false
nhanes_demographics %>%
  inner_join(nhanes_health, by = "SEQN") %>%
  select(SEQN, Age, Stroke)
```
:::

will include only IDs that are in the intersection of `nhanes_demographics` and `nhanes_health`.

### Outer Join

A full outer join:

::: {.panel-tabset}
# SQL

```{sql,connection=con}
SELECT SEQN, Age, Stroke
FROM nhanes_demographics
NATURAL FULL JOIN nhanes_health
```

# R

```{r}
#| edit: false
nhanes_demographics %>%
  full_join(nhanes_health, by = "SEQN") %>%
  select(SEQN, Age, Stroke)
```
:::

will include all IDs that are in either `nhanes_demographics` or `nhanes_health`. For patients that appear in `nhanes_demographics`, but not in `nhanes_health` (i.e., 93972 and 94069), null values will be inserted to `Stroke` (column in `nhanes_health`). Similarly, For patients that appear in `nhanes_health`, but not in `nhanes_demographics` (i.e., 102838, 102880, and 102947), null values will be inserted to `Age` (column in `nhanes_demographics`).

### Left outer join

A left outer join:

::: {.panel-tabset}
# SQL

```{sql,connection=con}
SELECT SEQN, Age, Stroke
FROM nhanes_demographics
NATURAL LEFT JOIN nhanes_health
```

# R

```{r}
#| edit: false
nhanes_demographics %>%
  left_join(nhanes_health, by = "SEQN") %>%
  select(SEQN, Age, Stroke)
```
:::

will include all IDs that are in the intersection (i.e., 93731, 93840, and 93920) plus those that are in `nhanes_demographics` only (i.e., 93972 and 94069) with null values inserted to `Stroke` (column in `nhanes_health`).

### Right outer join

A right outer join:

::: {.panel-tabset}
# SQL

```{sql,connection=con}
SELECT SEQN, Age, Stroke
FROM nhanes_demographics
NATURAL RIGHT JOIN nhanes_health
```

# R

```{r}
#| edit: false
nhanes_demographics %>%
  right_join(nhanes_health, by = "SEQN")%>%
  select(SEQN, Age, Stroke)
```
:::

will include all IDs that are in the intersection (i.e., 93731, 93840, and 93920) plus those that are in `nhanes_health` only (i.e., 102838, 102880, and 102947) with null values inserted to `Age` (column in `nhanes_demographics`).

## Return to Learning Hub Homepage

[Learning Hub Homepage](https://uoft-hmb-coding.github.io){.btn .btn-primary}